---
title: "Cross-ancestry fine-mapping of placental xQTL loci"
subtitle: "Fine-mapping report (Report 2 of 2) — SuSHiE credible sets, PIPs, and enrichment analyses"
author: "Bhattacharya Lab (analysis: S. Bresnahan; report generated with Biomni/Phylo)"
date: "`r format(Sys.Date())`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    self_contained: true
    df_print: paged
params:
  agg_dir: "data/extras"
  pips_path: "data/finemap/finemap_pips.tsv.gz"
  enrichment_path: "data/finemap/finemap_enrichment.tsv"
  geneset_enrichment_path: "data/finemap/finemap_geneset_enrichment.rds"
  data_dir: "data/results"
  module_path: "xqtl_report_functions.R"
  gene_map_path: "gene_map.tsv"
  fig_dir: "figures_finemap"
  table_dir: "tables_finemap"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.width = 8, fig.height = 5, dpi = 150)

suppressPackageStartupMessages({
  library(ggplot2)
  library(ggprism)
  library(dplyr)
  library(readr)
  library(tidyr)
  library(scales)
  library(data.table)
  library(patchwork)
})
source(params$module_path)   # report theme, MODALITY_LEVELS, loaders
theme_set(theme_prism() + theme(text = element_text(family = "Liberation Sans")))
dir.create(params$fig_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(params$table_dir, showWarnings = FALSE, recursive = TRUE)

gene_map <- read_tsv(params$gene_map_path, col_types = "cccc", progress = FALSE)
sym <- function(ids) gene_map$symbol[match(ids, gene_map$gene_id)]
symd <- function(ids) ifelse(is.na(sym(ids)) | sym(ids) == "", ids, sym(ids))
fmt <- function(x) format(x, big.mark = ",", scientific = FALSE, trim = TRUE)
extract_gene <- function(p) sub("\\..*$", "", sub("__.*$", "", p))

loci <- fread(file.path(params$agg_dir, "finemap_locus_summary.tsv"))
cs   <- fread(file.path(params$agg_dir, "finemap_credible_sets.tsv.gz"))
has_pips <- file.exists(params$pips_path)
has_enr  <- file.exists(params$enrichment_path)
has_gsenr <- file.exists(params$geneset_enrichment_path)

corr_cols <- grep("_est_corr$", names(cs), value = TRUE)
save_fig <- function(p, name, w = 8, h = 5) {
  print(p)   # ensure knitr embeds the figure even when save_fig is not the last expression
  ggsave(file.path(params$fig_dir, paste0(name, ".png")), p,
         width = w, height = h, dpi = 150, bg = "white")
  invisible(p)
}
```

# Overview

SuSHiE jointly fine-maps each significant xQTL locus across ancestry strata,
leveraging ancestry-specific LD to refine 95% credible sets and estimate
cross-ancestry effect correlations. This report summarizes run diagnostics,
credible-set properties, cross-ancestry concordance, top loci, and two
complementary enrichment analyses: functional-annotation enrichment of
high-PIP variants, and gene-set enrichment of high-confidence fine-mapped
genes. A literature review of the top fine-mapped genes lives in the
companion document `literature_finemap.md`.

```{r overview}
n_ok <- sum(loci$status == "ok", na.rm = TRUE)
n_conv <- sum(loci$converged == TRUE, na.rm = TRUE)
tibble(
  Metric = c("Loci attempted", "Loci fine-mapped", "Converged (min-tolerance)",
             "Loci with >=1 credible set", "Credible sets total",
             "Single-variant credible sets",
             "Loci fine-mapped in both ancestries"),
  Value = c(fmt(nrow(loci)), fmt(n_ok), fmt(n_conv),
            fmt(sum(loci$n_cs > 0, na.rm = TRUE)), fmt(nrow(cs)),
            fmt(sum(cs$n_variants == 1, na.rm = TRUE)),
            fmt(sum(loci$n_ancestries == 2, na.rm = TRUE)))
) |> knitr::kable()
```

# PIP diagnostics

```{r fig_maxpip, fig.cap="Maximum joint PIP per locus."}
p <- ggplot(loci |> filter(!is.na(max_pip)), aes(x = max_pip)) +
  geom_histogram(bins = 50, fill = "#0279EE", color = "white") +
  labs(x = "Maximum joint PIP per locus", y = "Loci")
save_fig(p, "fig_fm_maxpip")
```

```{r fig_pipdist, fig.cap="Distribution of joint PIPs across all fine-mapped variants (log scale).", eval=has_pips}
if (has_pips) {
  pips_piponly <- fread(params$pips_path, select = c("pip_all"))
  p <- ggplot(pips_piponly, aes(x = pip_all)) +
    geom_histogram(bins = 60, fill = "#75A025", color = "white") +
    scale_y_log10() +
    labs(x = "Joint PIP", y = "Variants (log10)")
  save_fig(p, "fig_fm_pipdist")
  rm(pips_piponly); invisible(gc(verbose = FALSE))
}
```

# Credible sets

```{r fig_cssize, fig.cap="Credible-set size distribution (number of variants per 95% CS)."}
if (nrow(cs) > 0) {
  p <- ggplot(cs, aes(x = n_variants)) +
    geom_histogram(bins = 40, fill = "#FF9400", color = "white") +
    scale_x_log10() +
    labs(x = "Variants per credible set (log10)", y = "Credible sets")
  save_fig(p, "fig_fm_cs_size")
} else {
  cat("No credible sets passed purity pruning.\n")
}
```

```{r fig_csperlocus, fig.cap="Credible sets per locus."}
if (nrow(cs) > 0) {
  p <- cs |> count(modality, phenotype_id) |>
    ggplot(aes(x = factor(n))) +
    geom_bar(fill = "#0279EE") +
    labs(x = "Credible sets per locus", y = "Loci")
  save_fig(p, "fig_fm_cs_per_locus")
}
```

```{r fig_csleadpip, fig.cap="Lead-variant PIP per credible set. The bimodal distribution separates confidently resolved signals (lead PIP near 1) from loci where the causal variant remains uncertain."}
p <- ggplot(cs, aes(x = lead_pip)) +
  geom_histogram(bins = 50, fill = "#75A025", color = "white") +
  labs(x = "Credible-set lead PIP", y = "Credible sets")
save_fig(p, "fig_fm_cs_leadpip")
```

## Credible sets versus conditional signals

*DevBrain Fig. 1I analog: the number of SuSHiE credible sets per locus
versus the number of conditionally independent signals at the same
phenotype (maximum rank across ancestries in the stepwise scans).*

```{r fig-cs-vs-conditional, fig.width=7.5, fig.height=5.5}
independent <- load_independent(params$data_dir)
cond_counts <- independent %>%
  filter(layer == "ungrouped" | (layer == "grouped" & modality == "expression")) %>%
  group_by(phenotype_id) %>%
  summarise(n_cond = max(rank), .groups = "drop")
cs_join <- loci %>%
  filter(!is.na(n_cs_out)) %>%
  inner_join(cond_counts, by = "phenotype_id")
cs_pts <- cs_join %>% count(n_cond, n_cs_out, name = "n_loci")
rho_cs <- if (nrow(cs_join) > 5)
  cor(cs_join$n_cond, cs_join$n_cs_out, method = "spearman") else NA_real_
p_cs <- ggplot(cs_pts, aes(n_cond, n_cs_out)) +
  geom_point(aes(size = n_loci), alpha = 0.55, color = "#0279EE") +
  scale_size_area(max_size = 12, name = "loci") +
  scale_x_continuous(breaks = scales::breaks_width(1)) +
  scale_y_continuous(breaks = scales::breaks_width(1)) +
  annotate("label", x = Inf, y = -Inf, hjust = 1.05, vjust = -0.8,
           fill = "white", alpha = 0.85, label.size = 0, size = 3.5,
           label = sprintf("Spearman rho = %.2f\n(n = %s loci)", rho_cs,
                           fmt(nrow(cs_join)))) +
  labs(x = "conditionally independent signals (max across ancestries)",
       y = "SuSHiE credible sets per locus") +
  theme_report()
print(p_cs)
ggsave(file.path(params$fig_dir, "fig_cs_vs_conditional.png"), p_cs,
       width = 7.5, height = 5.5, dpi = 150)
```

`r sprintf("%s of %s fine-mapped loci with credible sets matched a stepwise-conditional scan; the credible-set count tracks the conditional-signal count (Spearman rho = %.2f), cross-validating the two independence estimates.", fmt(nrow(cs_join)), fmt(sum(!is.na(loci$n_cs_out))), rho_cs)`

# Cross-ancestry concordance

```{r fig_rho, fig.cap="Cross-ancestry effect correlation (rho) per credible set. Values near 1 indicate shared causal effects; low/negative values indicate ancestry-specific effects."}
if (nrow(cs) > 0 && length(corr_cols) > 0) {
  rho_long <- cs |>
    select(modality, phenotype_id, cs_index, all_of(corr_cols)) |>
    pivot_longer(all_of(corr_cols), names_to = "pair", values_to = "rho") |>
    filter(!is.na(rho))
  p <- ggplot(rho_long, aes(x = rho)) +
    geom_histogram(bins = 40, fill = "#75A025", color = "white") +
    labs(x = "Cross-ancestry effect correlation (rho)", y = "Credible sets")
  save_fig(p, "fig_fm_rho")
} else {
  cat("No cross-ancestry correlations (single-ancestry loci only).\n")
}
```

`r if ("EAS_EUR_est_corr" %in% names(cs)) sprintf("The cross-ancestry effect-correlation distribution (n = %s credible sets fine-mapped in both ancestries) centers at %.2f, quantifying how shared the causal effects are between strata.", fmt(sum(!is.na(cs$EAS_EUR_est_corr))), median(cs$EAS_EUR_est_corr, na.rm = TRUE)) else ""`

```{r fig_weights, fig.cap="Per-ancestry posterior effect weights for credible-set variants. Concordant effects lie along the diagonal.", eval=has_pips}
if (has_pips) {
  pips_w <- fread(params$pips_path,
                  select = c("modality", "phenotype_id", "cs_index",
                             "EUR_effect_weight", "EAS_effect_weight"))
  anc_cols <- grep("_effect_weight$", names(pips_w), value = TRUE)
  anc_labels <- sub("_effect_weight$", "", anc_cols)
  cs_vars <- pips_w |> filter(!is.na(cs_index))
  n_cs_levels <- length(unique(cs_vars$cs_index))
  okabe_ito <- c("#E69F00", "#0072B2", "#009E73", "#CC79A7", "#D55E00",
                 "#56B4E9", "#F0E442", "#000000")
  p <- ggplot(cs_vars,
              aes(x = .data[[anc_cols[1]]], y = .data[[anc_cols[2]]],
                  color = factor(cs_index))) +
    geom_point(alpha = 0.7, size = 1.6) +
    scale_color_manual(values = colorRampPalette(okabe_ito)(n_cs_levels)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed",
                color = "grey50") +
    labs(x = paste(anc_labels[1], "posterior effect weight"),
         y = paste(anc_labels[2], "posterior effect weight"),
         color = "CS index") +
    theme(legend.position = "none")
  save_fig(p, "fig_fm_effect_concordance")
  rm(pips_w); invisible(gc(verbose = FALSE))
}
```

# Top loci

```{r top-loci-table}
cs[, gene_id := extract_gene(phenotype_id)]
# gene-level recurrence: number of modalities with a high-confidence locus
gene_rec <- cs[lead_pip >= 0.9, .(n_mod = length(unique(modality))),
               by = .(gene_id)]
top_loci_tbl <- cs |>
  group_by(modality, phenotype_id, gene_id) |>
  summarise(best_lead_pip = max(lead_pip), n_cs = n(),
            min_cs_size = min(n_variants), .groups = "drop") |>
  left_join(gene_rec, by = "gene_id") |>
  mutate(gene_symbol = symd(gene_id)) |>
  arrange(desc(best_lead_pip), min_cs_size, desc(n_mod))
write_tsv(top_loci_tbl, file.path(params$table_dir, "finemap_top_loci.tsv"))
knitr::kable(top_loci_tbl |>
               select(modality, gene_symbol, phenotype_id, best_lead_pip,
                      n_cs, min_cs_size, n_mod) |>
               head(30),
             digits = 3,
             caption = "Top 30 fine-mapped loci (best credible-set lead PIP, ties broken by smallest credible set, then cross-modality recurrence of the gene; full table in tables_finemap/finemap_top_loci.tsv).")
```

```{r fig_locus_tracks, fig.cap="Joint PIP by position for the six loci with the highest lead-variant PIP. Points in a credible set are colored by CS index.", eval=has_pips}
if (has_pips && nrow(cs) > 0) {
  top_loci <- cs |> arrange(desc(lead_pip), n_variants) |>
    distinct(modality, phenotype_id, .keep_all = TRUE) |>
    slice_head(n = 6) |> select(modality, phenotype_id)
  pips_tracks <- fread(params$pips_path,
                       select = c("modality", "phenotype_id", "pos",
                                  "pip_all", "cs_index"))
  tracks <- pips_tracks |>
    inner_join(top_loci, by = c("modality", "phenotype_id")) |>
    mutate(gene = symd(extract_gene(phenotype_id)),
           locus = paste(modality, gene, sep = " / "),
           in_cs = ifelse(is.na(cs_index), "not in CS",
                          paste0("CS", cs_index))) |>
    arrange(in_cs == "not in CS")   # draw CS points last (on top)
  cs_levels <- sort(setdiff(unique(tracks$in_cs), "not in CS"))
  okabe_ito <- c("#E69F00", "#0072B2", "#009E73", "#CC79A7", "#D55E00",
                 "#56B4E9", "#F0E442", "#000000")
  cs_pal <- setNames(c("grey75", colorRampPalette(okabe_ito)(length(cs_levels))),
                     c("not in CS", cs_levels))
  p <- ggplot(tracks, aes(x = pos / 1e6, y = pip_all, color = in_cs,
                          size = in_cs == "not in CS")) +
    geom_point(alpha = 0.8) +
    scale_size_manual(values = c(`TRUE` = 0.7, `FALSE` = 2.0), guide = "none") +
    scale_color_manual(values = cs_pal) +
    facet_wrap(~locus, scales = "free_x", ncol = 2) +
    labs(x = "Position (Mb)", y = "Joint PIP", color = NULL) +
    theme(legend.position = "bottom")
  save_fig(p, "fig_fm_locus_tracks", w = 9, h = 7)
  rm(pips_tracks); invisible(gc(verbose = FALSE))
}
```

# Functional-annotation enrichment of high-PIP variants

High-PIP variants (PIP >= 0.9) tested for enrichment against fastVEP
consequence classes, ENCODE SCREEN cCRE classes, and placenta open-chromatin
peaks; background = all other fine-mapped variants. Odds ratios and 95% CIs
are recomputed from the 2x2 cell counts; FDR is Benjamini-Hochberg across
classes.

```{r fig_enrichment, fig.cap="Annotation enrichment of high-PIP variants (Fisher exact odds ratio, 95% CI)."}
if (has_enr) {
  enr <- read_tsv(params$enrichment_path, col_types = cols(), progress = FALSE)
  ci <- lapply(seq_len(nrow(enr)), function(i) {
    m <- matrix(c(enr$cell_a_high_in[i], enr$cell_b_high_out[i],
                  enr$cell_c_bg_in[i], enr$cell_d_bg_out[i]), nrow = 2,
                byrow = TRUE)
    ft <- fisher.test(m)
    tibble(or = unname(ft$estimate), lo = ft$conf.int[1],
           hi = ft$conf.int[2])
  }) |> bind_rows()
  enr_plot <- enr |> bind_cols(ci) |>
    mutate(annotation = reorder(annotation, or))
  write_tsv(enr_plot, file.path(params$table_dir,
                                "pip_annotation_enrichment_with_ci.tsv"))
  p <- enr_plot |>
    ggplot(aes(x = annotation, y = or)) +
    geom_hline(yintercept = 1, linetype = "dashed", color = "grey50") +
    geom_pointrange(aes(ymin = lo, ymax = pmin(hi, quantile(hi[is.finite(hi)], 0.95) * 3)),
                    color = "#0279EE") +
    coord_flip() +
    scale_y_log10() +
    labs(x = NULL, y = "Odds ratio (log10, 95% CI; upper CI capped for display)")
  save_fig(p, "fig_fm_enrichment", w = 8, h = 6)
  knitr::kable(enr_plot |>
                 arrange(desc(or)) |>
                 select(annotation, n_variants, n_high_pip, odds_ratio,
                        pip_weighted_enrichment, fisher_fdr) |>
                 mutate(across(where(is.numeric), ~ signif(.x, 3))),
               caption = "High-PIP variant annotation enrichment (sorted by odds ratio).")
} else {
  cat("Enrichment table not found; run 34_pip_annotation_enrichment.py.\n")
}
```

# Gene-set enrichment of high-confidence fine-mapped genes

*Per-modality over-representation analysis of genes whose best credible set
resolves to a high-confidence causal variant (lead PIP >= 0.9), against GO
(BP/MF/CC), KEGG, MSigDB Hallmark, and the curated pregnancy/placenta
collection. Primary background: all fine-mapped genes of the same modality
(tests what distinguishes confidently resolved loci among significant ones);
sensitivity background: the HPLRv2 reference universe. BH FDR within each
modality x database x background. Computed by `run_finemap_enrichment.R`;
results read from cache.*

```{r geneset-enrichment-load}
if (has_gsenr) {
  gsenr <- readRDS(params$geneset_enrichment_path)
  gs_all <- gsenr$results
  gs_sizes <- gsenr$sizes
  gs_sig <- gs_all[p.adjust < 0.05]
  knitr::kable(gs_sizes,
               caption = "Per-modality gene counts: fine-mapped (universe) and high-confidence (lead PIP >= 0.9).")
} else {
  cat("Gene-set enrichment cache not found; run run_finemap_enrichment.R.\n")
}
```

```{r geneset-sig-counts}
if (has_gsenr) {
  counts <- gs_sig[background == "finemapped", .N, by = .(modality, db)]
  counts_wide <- dcast(counts, modality ~ db, value.var = "N", fill = 0)
  knitr::kable(counts_wide,
               caption = "Number of significant terms (FDR < 0.05) per modality, fine-mapped background")
}
```

```{r fig-gsenrich-go, fig.width=11, fig.height=8, fig.cap="Top GO Biological Process terms per modality (fine-mapped background)."}
if (has_gsenr) {
  gobp <- gs_all[db == "GO_BP" & background == "finemapped"]
  top5 <- gobp[order(p.adjust), head(.SD, 5), by = .(modality)]
  if (nrow(top5) > 0) {
    top5[, neglog := -log10(p.adjust)]
    top5[, term := substr(Description, 1, 45)]
    top5[, term_ord := reorder(term, neglog, max)]
    p_go <- ggplot(top5, aes(x = modality, y = term_ord)) +
      geom_point(aes(color = neglog, size = Count)) +
      scale_color_gradient(low = "#74add1", high = "#d73027",
                           name = expression(-log[10](p[adj]))) +
      scale_size_continuous(name = "Genes", range = c(1.5, 6)) +
      labs(x = NULL, y = NULL) +
      theme_bw(base_family = "Liberation Sans") +
      theme(axis.text.x = element_text(size = 8, angle = 30, hjust = 1),
            axis.text.y = element_text(size = 7),
            legend.text = element_text(size = 7),
            legend.title = element_text(size = 7),
            panel.grid.minor = element_blank())
    print(p_go)
    ggsave(file.path(params$fig_dir, "fig_fm_gsenrich_go.png"), p_go,
           width = 11, height = 8, dpi = 150, bg = "white")
  }
}
```

```{r gsenrich-tables}
if (has_gsenr) {
  sig_tab <- gs_sig[background == "finemapped",
                    .(modality, db, Description, GeneRatio, BgRatio,
                      p.adjust = signif(p.adjust, 3))]
  knitr::kable(sig_tab[order(db, modality, p.adjust)],
               caption = "Significant terms (FDR < 0.05), fine-mapped background")
}
```

```{r fig-gsenrich-pregnancy, fig.width=9, fig.height=6, fig.cap="Pregnancy/placenta gene-set enrichment of high-confidence fine-mapped genes (rings: FDR < 0.05)."}
if (has_gsenr) {
  pr <- gs_all[db == "PREGNANCY"]
  if (nrow(pr) > 0) {
    pr[, neglog := -log10(p.adjust)]
    pr[, set := factor(Description,
                       levels = c("PE_GWAS_candidates", "STB_syncytialization_markers",
                                  "HPA_placenta_enriched", "MSigDB_SU_PLACENTA",
                                  "WP_iron_metabolism_in_placenta"))]
    pr[, is_sig := p.adjust < 0.05]
    p_pr <- ggplot(pr, aes(x = modality, y = set)) +
      geom_point(aes(color = neglog, size = Count)) +
      geom_point(data = pr[is_sig == TRUE], shape = 21, color = "black",
                 stroke = 1.1, aes(size = Count), fill = NA) +
      scale_color_gradient(low = "#c7e9c0", high = "#006d2c",
                           name = expression(-log[10](p[adj]))) +
      scale_size_continuous(name = "Overlap", range = c(1, 7)) +
      facet_wrap(~ background, ncol = 1,
                 labeller = labeller(
                   background = c(finemapped = "Fine-mapped genes background",
                                  hplrv2 = "HPLRv2 reference background"))) +
      labs(x = NULL, y = NULL) +
      theme_bw(base_family = "Liberation Sans") +
      theme(axis.text.x = element_text(size = 8, angle = 30, hjust = 1),
            axis.text.y = element_text(size = 8),
            strip.text = element_text(face = "bold", size = 9),
            legend.text = element_text(size = 7),
            legend.title = element_text(size = 7),
            panel.grid.minor = element_blank()) +
      scale_y_discrete(drop = FALSE)
    print(p_pr)
    ggsave(file.path(params$fig_dir, "fig_fm_gsenrich_pregnancy.png"), p_pr,
           width = 9, height = 6, dpi = 150, bg = "white")
  }
}
```

```{r gsenrich-background-sensitivity}
if (has_gsenr) {
  sig_wide <- dcast(gs_sig[, .(modality, db, ID, Description, background)],
                    modality + db + ID + Description ~ background,
                    value.var = "background", fun.aggregate = length)
  for (cl in c("finemapped", "hplrv2"))
    if (!cl %in% names(sig_wide)) sig_wide[, (cl) := 0L]
  sig_wide[, concord := fifelse(finemapped > 0 & hplrv2 > 0, "both",
                                fifelse(finemapped > 0, "finemapped-only",
                                        "hplrv2-only"))]
  knitr::kable(sig_wide[, .N, by = concord],
               caption = "Concordance of significant term-modalities between backgrounds")
}
```

# Run diagnostics

```{r diagnostics}
loci |> count(status) |> knitr::kable(col.names = c("Status", "Loci"))
if (nrow(loci |> filter(status == "ok")) > 0) {
  p <- loci |> filter(status == "ok") |>
    ggplot(aes(x = wall_sec)) +
    geom_histogram(bins = 40, fill = "#FF9400", color = "white") +
    labs(x = "Per-locus wall time (s)", y = "Loci")
  save_fig(p, "fig_fm_walltime")
}
```

# Methods summary

- **Fine-mapping**: SuSHiE joint cross-ancestry fine-mapping of the
  FDR-significant phenotype-level discovery set (ungrouped layer +
  expression primary scan), per-locus 95% credible sets with purity pruning,
  and cross-ancestry effect correlations (pipeline scripts 31-33).
- **Variant-annotation enrichment**: high-PIP variants (PIP >= 0.9) vs all
  other fine-mapped variants; fastVEP consequence classes, ENCODE SCREEN
  cCRE classes, placenta open-chromatin peaks; Fisher exact + PIP-weighted
  enrichment, BH FDR (script 34).
- **Gene-set enrichment**: per modality, genes with best-CS lead PIP >= 0.9
  vs (a) all fine-mapped genes of the modality and (b) the HPLRv2 reference;
  clusterProfiler `enrichGO`/`enrichKEGG`/`enricher` (Hallmark via msigdbr;
  curated pregnancy/placenta collection identical to the xQTL-mapping
  report); BH FDR within modality x database x background
  (`run_finemap_enrichment.R`).
- **Companion literature review**: `literature_finemap.md` reviews the top
  fine-mapped genes against the published literature (pregnancy/birth
  outcomes, intragenomic conflict, placental biology).

# Session info

```{r session-info}
sessionInfo()
```
