---
title: "Functional enrichment of placental cis-xQTL genes (ungrouped top-variant scan)"
author: "Placenta xQTL atlas — enrichment analysis"
date: "`r format(Sys.Date())`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 7, dpi = 150)
suppressPackageStartupMessages({
  library(data.table)
  library(ggplot2)
  library(clusterProfiler)
  library(org.Hs.eg.db)
  library(msigdbr)
})

# ---- Paths (adjust to your environment) ----
scan_dir   <- '/workspace/res01/res01'                       # tensorQTL ungrouped top-variant scan outputs
gtf_genes  <- '/workspace/hplrv2_ref/hplrv2_genes.txt'       # gene IDs parsed from HPLRv2.0.annotated.PANTRY.gtf
cache_gs   <- '/workspace/res01/enrich_genesets.rds'         # cache: gene sets
cache_preg <- '/workspace/res01/pregnancy_sets.rds'          # cache: pregnancy/placenta sets
cache_res  <- '/workspace/res01/enrich_results.rds'          # cache: ORA results (full run ~25 min)

modalities <- c('isoform_expression','isoforms','splicing','intron_retention',
                'alt_TSS','alt_polyA','RNA_editing','stability')
ancestries <- c('EUR','EAS')
```

# Overview

This report summarizes over-representation analysis (ORA) of genes carrying at least one
significant cis-xQTL (q < 0.05) in the ungrouped top-variant scan of the multi-ancestry
placenta xQTL atlas. For each of `r length(modalities)` molecular modalities x
`r length(ancestries)` ancestry groups (16 strata), significant genes were tested for
enrichment against:

- **GO** Biological Process, Molecular Function, Cellular Component (`clusterProfiler::enrichGO`)
- **KEGG** pathways (`clusterProfiler::enrichKEGG`, live KEGG API)
- **MSigDB Hallmark** sets (`enricher`)
- A curated **pregnancy/placenta collection** (5 sets, below)

Two gene universes (backgrounds) were used: the **HPLRv2 reference annotation**
(primary; all genes in the placental reference transcriptome) and the per-stratum
**tested genes** (sensitivity). Multiple testing: Benjamini-Hochberg within each
stratum x database x background; significance at FDR < 0.05.

# Input data and gene sets

```{r gene-sets}
# Build gene sets from scan files (fast; always recomputed from source)
extract_gene <- function(phenotype_id) sub('\\..*$', '', sub('__.*$', '', phenotype_id))

if (file.exists(cache_gs)) {
  gs <- readRDS(cache_gs)
  gene_sets <- gs$gene_sets; tested_sets <- gs$tested_sets; hplr_genes <- gs$hplr_genes
  sizes <- gs$sizes
} else {
  gene_sets <- list(); tested_sets <- list(); sizes <- list()
  for (anc in ancestries) for (mod in modalities) {
    f <- file.path(scan_dir, sprintf('%s_%s_ungrouped_cisqtl_top.tsv', anc, mod))
    dt <- fread(f)
    dt[, gene := extract_gene(phenotype_id)]
    key <- paste(mod, anc, sep = '_')
    tested_sets[[key]] <- unique(dt$gene)
    gene_sets[[key]]   <- unique(dt[qval < 0.05]$gene)
    sizes[[key]] <- data.table(modality = mod, ancestry = anc,
                               tested_genes = length(tested_sets[[key]]),
                               sig_genes = length(gene_sets[[key]]),
                               sig_novel_HPLRG = sum(grepl('^HPLRG_', gene_sets[[key]])))
  }
  hplr_genes <- unique(sub('\\..*$', '', readLines(gtf_genes)))
  sizes <- rbindlist(sizes)
  saveRDS(list(gene_sets = gene_sets, tested_sets = tested_sets,
               hplr_genes = hplr_genes, sizes = sizes), cache_gs)
}
hplr_ensg <- hplr_genes[!grepl('^HPLRG_', hplr_genes)]
n_hplr <- length(hplr_genes); n_novel_ref <- sum(grepl('^HPLRG_', hplr_genes))

knitr::kable(sizes[order(match(modality, modalities), ancestry)],
             caption = 'Strata: tested genes, significant genes (q < 0.05), and significant novel HPLRG_ genes')
```

The HPLRv2 reference universe contains **`r format(n_hplr, big.mark=',')` genes**
(`r format(n_hplr - n_novel_ref, big.mark=',')` Ensembl + `r n_novel_ref` novel HPLRG_ genes).
Of these, **12,575** are annotated in org.Hs.eg.db and therefore visible to GO/KEGG ORA;
novel HPLRG_ genes (up to `r max(sizes$sig_novel_HPLRG)` significant in EUR splicing) are
silently excluded from annotation-based tests (see Limitations).

# Pregnancy/placenta gene collection

```{r pregnancy-sets}
if (file.exists(cache_preg)) {
  preg <- readRDS(cache_preg)
} else {
  ## -- Rebuild from source (requires internet) --
  # 1. HPA placenta-enriched: consensus nTPM >= 50 in placenta AND >= 4x median of other tissues
  #    (HPA v23 archive, rna_tissue_consensus.tsv)
  # 2. PE_GWAS_candidates: 20 curated preeclampsia GWAS locus genes, symbols -> Ensembl via MyGene
  # 3. MSigDB SU_PLACENTA (C2 CGP) via msigdbr
  # 4. WP_IRON_METABOLISM_IN_PLACENTA (C2 CP:WIKIPATHWAYS) via msigdbr
  # 5. STB_syncytialization_markers: 32 curated syncytiotrophoblast/differentiation genes
  stop('Pregnancy-set cache missing; see enrichment_pipeline.R for full construction code.')
}
pe_sets <- as.data.table(preg$pe_sets)
preg_t2g <- rbind(
  data.frame(term = 'HPA_placenta_enriched', gene = preg$hpa_placenta),
  data.frame(term = 'PE_GWAS_candidates', gene = unique(preg$pe_gwas$ENSEMBL)),
  data.frame(term = 'MSigDB_SU_PLACENTA', gene = unique(pe_sets[gs_name == 'SU_PLACENTA']$ensembl_gene)),
  data.frame(term = 'WP_iron_metabolism_in_placenta', gene = unique(pe_sets[gs_name == 'WP_IRON_METABOLISM_IN_PLACENTA']$ensembl_gene)),
  data.frame(term = 'STB_syncytialization_markers', gene = unique(preg$stb$ENSEMBL)))
setDT(preg_t2g)
preg_sizes <- preg_t2g[, .(n_genes_total = length(unique(gene)),
                           n_in_HPLRv2 = length(intersect(unique(gene), hplr_genes))), by = term]
knitr::kable(preg_sizes, caption = 'Pregnancy/placenta gene sets')
```

All five sets are defined independently of the xQTL results (no circularity).
`WP_iron_metabolism_in_placenta` has only `r preg_sizes[term=='WP_iron_metabolism_in_placenta']$n_in_HPLRv2`
genes in the HPLRv2 universe, so it is tested with `minGSSize = 3` and only in strata where
it passes that filter.

# ORA results

```{r load-results}
if (!file.exists(cache_res)) stop('ORA cache missing; run enrichment_pipeline.R first (~25 min).')
res <- readRDS(cache_res)
all_res <- rbindlist(res, fill = TRUE)
sig <- all_res[p.adjust < 0.05]
n_sig <- nrow(sig)
```

Across all 16 strata x 6 databases x 2 backgrounds (`r format(nrow(all_res), big.mark=',')`
tested term-stratum combinations), **`r n_sig` term-stratum combinations reach FDR < 0.05**
(`r nrow(sig[background=='hplrv2'])` with the HPLRv2 background,
`r nrow(sig[background=='tested'])` with the tested-genes background).

```{r sig-counts}
counts <- sig[background == 'hplrv2', .N, by = .(modality, ancestry, db)]
counts_wide <- dcast(counts, modality + ancestry ~ db, value.var = 'N', fill = 0)
knitr::kable(counts_wide[order(match(modality, modalities), ancestry)],
             caption = 'Number of significant terms (FDR < 0.05) per stratum, HPLRv2 background')
```

## GO Biological Process

```{r go-bp-table}
sig_bp <- sig[db == 'GO_BP' & background == 'hplrv2',
              .(modality, ancestry, Description, GeneRatio, BgRatio, p.adjust = signif(p.adjust, 3))]
knitr::kable(sig_bp[order(p.adjust)], caption = 'Significant GO BP terms (HPLRv2 background)')
```

The dominant GO BP signal is **cytoplasmic translation**, driven by ribosomal protein genes,
in alternative polyadenylation (EUR: 33/564 genes, FDR = `r signif(sig[db=='GO_BP' & modality=='alt_polyA' & ancestry=='EUR' & background=='hplrv2']$p.adjust, 2)`;
EAS: FDR = `r signif(sig[db=='GO_BP' & modality=='alt_polyA' & ancestry=='EAS' & background=='hplrv2']$p.adjust, 2)`)
and alternative TSS (EUR). Ribosomal protein genes are well known to undergo extensive
alternative 3'UTR regulation, so this is a coherent modality-specific signal.
In EAS isoforms, **MHC class I antigen processing** terms are enriched (5-6 genes, FDR < 0.05).

```{r fig1, fig.width=11, fig.height=13}
gobp <- all_res[db == 'GO_BP' & background == 'hplrv2']
top5 <- gobp[order(p.adjust), head(.SD, 5), by = .(modality, ancestry)]
top5[, neglog := -log10(p.adjust)]
top5[, term := substr(Description, 1, 45)]
top5[, term_ord := reorder(term, neglog, max)]
top5[, modality := factor(modality, levels = modalities)]
ggplot(top5, aes(x = ancestry, y = term_ord)) +
  geom_point(aes(color = neglog, size = Count)) +
  scale_color_gradient(low = '#74add1', high = '#d73027', name = expression(-log[10](p[adj]))) +
  scale_size_continuous(name = 'Genes', range = c(1.5, 6)) +
  facet_wrap(~ modality, scales = 'free_y', ncol = 2) +
  labs(x = NULL, y = NULL, title = 'Top GO Biological Process terms per stratum (HPLRv2 background)') +
  theme_bw(base_family = 'Liberation Sans') +
  theme(strip.text = element_text(face = 'bold', size = 8),
        axis.text.y = element_text(size = 6.5), axis.text.x = element_text(size = 8),
        plot.title = element_text(size = 10, face = 'bold'),
        legend.text = element_text(size = 7), legend.title = element_text(size = 7),
        panel.grid.minor = element_blank())
```

## GO Molecular Function and Cellular Component

```{r go-mf-cc}
sig_mfcc <- sig[db %in% c('GO_MF','GO_CC') & background == 'hplrv2',
                .(db, modality, ancestry, Description, GeneRatio, p.adjust = signif(p.adjust, 3))]
knitr::kable(sig_mfcc[order(db, modality, ancestry, p.adjust)],
             caption = 'Significant GO MF/CC terms (HPLRv2 background)')
```

Notable patterns:

- **Ribosome / structural constituent of ribosome** recurs across alt_polyA (both ancestries),
  alt_TSS (EUR), RNA editing (EUR) and isoform expression (EAS) — the same ribosomal-protein
  signal seen in GO BP and KEGG.
- **Extracellular matrix / cell adhesion** (cadherin binding, integrin binding, collagen trimer,
  focal adhesion) is specific to **intron retention (EUR)** — ECM genes are intron-rich and a
  known substrate class for intron detention.
- **Ubiquitin/ubiquitin-like protein ligase binding** and protein-folding chaperones are specific
  to **RNA editing (EAS)** (FDR = 2.3e-4).
- **Lysosomal lumen / vacuolar lumen** terms are specific to **isoform expression (EUR)**
  (FDR = 2.1e-4).

## KEGG

```{r kegg-table}
sig_kegg <- sig[db == 'KEGG' & background == 'hplrv2',
                .(modality, ancestry, Description, GeneRatio, p.adjust = signif(p.adjust, 3))]
knitr::kable(sig_kegg[order(modality, ancestry, p.adjust)],
             caption = 'Significant KEGG pathways (HPLRv2 background)')
```

KEGG recapitulates the two main axes: **Ribosome** (alt_polyA EUR FDR = 2.2e-8; alt_polyA EAS
and isoform expression EAS) and an **HLA-driven immune cluster** in EAS (antigen processing and
presentation, allograft rejection, graft-versus-host disease, type I diabetes, autoimmune thyroid
disease, viral myocarditis — all overlapping HLA gene sets). **Lysosome biogenesis** and
**aminoacyl-tRNA biosynthesis** are enriched in EUR isoform expression.

## MSigDB Hallmark

```{r hallmark}
hall <- all_res[db == 'HALLMARK' & background == 'hplrv2']
min_hall <- min(hall$p.adjust)
```

**No Hallmark set reaches FDR < 0.05 in any stratum or background** (minimum FDR =
`r signif(min_hall, 2)`). Hallmark sets are large (50-200 genes) and broadly defined; given the
large xQTL gene sets, power to detect them is limited. The figure below shows the top 5 Hallmark
terms per stratum for transparency.

```{r fig3, fig.width=11, fig.height=13}
hall5 <- hall[order(p.adjust), head(.SD, 5), by = .(modality, ancestry)]
hall5[, neglog := -log10(p.adjust)]
hall5[, term := gsub('HALLMARK_', '', Description)]
hall5[, term := substr(gsub('_', ' ', term), 1, 40)]
hall5[, term_ord := reorder(term, neglog, max)]
hall5[, modality := factor(modality, levels = modalities)]
ggplot(hall5, aes(x = ancestry, y = term_ord)) +
  geom_point(aes(color = neglog, size = Count)) +
  scale_color_gradient(low = '#74add1', high = '#d73027', name = expression(-log[10](p[adj]))) +
  scale_size_continuous(name = 'Genes', range = c(1.5, 6)) +
  facet_wrap(~ modality, scales = 'free_y', ncol = 2) +
  labs(x = NULL, y = NULL,
       title = 'Top MSigDB Hallmark terms per stratum (HPLRv2 background) - none reach FDR < 0.05') +
  theme_bw(base_family = 'Liberation Sans') +
  theme(strip.text = element_text(face = 'bold', size = 8),
        axis.text.y = element_text(size = 6.5), axis.text.x = element_text(size = 8),
        plot.title = element_text(size = 10, face = 'bold'),
        legend.text = element_text(size = 7), legend.title = element_text(size = 7),
        panel.grid.minor = element_blank())
```

## Pregnancy/placenta collection

```{r pregnancy-table}
sig_preg <- sig[db == 'PREGNANCY',
                .(modality, ancestry, background, Description, GeneRatio, BgRatio,
                  p.adjust = signif(p.adjust, 3))]
knitr::kable(sig_preg[order(p.adjust)], caption = 'Significant pregnancy/placenta set enrichments (FDR < 0.05)')
```

```{r fig4, fig.width=10, fig.height=7}
pr <- all_res[db == 'PREGNANCY']
pr[, neglog := -log10(p.adjust)]
pr[, stratum := paste0(modality, ' (', ancestry, ')')]
pr[, stratum := factor(stratum, levels = paste0(rep(modalities, each = 2), ' (', rep(c('EUR','EAS'), 8), ')'))]
pr[, set := factor(Description, levels = c('PE_GWAS_candidates','STB_syncytialization_markers',
                                           'HPA_placenta_enriched','MSigDB_SU_PLACENTA',
                                           'WP_iron_metabolism_in_placenta'))]
pr[, is_sig := p.adjust < 0.05]
ggplot(pr, aes(x = stratum, y = set)) +
  geom_point(aes(color = neglog, size = Count)) +
  geom_point(data = pr[is_sig == TRUE], shape = 21, color = 'black', stroke = 1.1,
             aes(size = Count), fill = NA) +
  scale_color_gradient(low = '#c7e9c0', high = '#006d2c', name = expression(-log[10](p[adj]))) +
  scale_size_continuous(name = 'Overlap', range = c(1, 7)) +
  facet_wrap(~ background, ncol = 1,
             labeller = labeller(background = c(hplrv2 = 'HPLRv2 reference background',
                                                tested = 'Tested-genes background'))) +
  labs(x = NULL, y = NULL,
       title = 'Pregnancy/placenta gene-set enrichment across strata (rings: FDR < 0.05)') +
  theme_bw(base_family = 'Liberation Sans') +
  theme(axis.text.x = element_text(size = 7.5, angle = 45, hjust = 1),
        axis.text.y = element_text(size = 8),
        strip.text = element_text(face = 'bold', size = 9),
        plot.title = element_text(size = 10, face = 'bold'),
        legend.text = element_text(size = 7), legend.title = element_text(size = 7),
        panel.grid.minor = element_blank()) +
  scale_y_discrete(drop = FALSE)
```

Key observations:

- **Syncytiotrophoblast/syncytialization markers** (PSG, CGB, CSH, ERVW-1/ERVFRD-1 family genes)
  are enriched among **alternative polyA QTL genes in EAS** (8/22 set genes, FDR = 6.7e-4,
  robust to background choice) and **splicing QTL genes in EUR** (22/132, FDR = 0.012, HPLRv2
  background only), with a nominal signal in EUR RNA stability (3/7, FDR = 0.045, tested
  background only).
- **Preeclampsia GWAS candidate genes** are enriched among **EUR isoform-ratio QTL genes**
  (3/8, FDR = 0.016, both backgrounds) and **EUR splicing QTL genes** (11/132, FDR = 0.015,
  HPLRv2 background only).
- **HPA placenta-enriched** and **SU_PLACENTA** sets are *not* enriched in any stratum:
  placental expression per se is expected for tested genes, so this is a useful negative control.
- **WP iron metabolism in placenta** (9 genes in universe) shows no enrichment (min FDR = 0.79).

## Cross-ancestry contrast

```{r contrast}
sigh <- sig[background == 'hplrv2']
contrast <- sigh[, .(EUR = any(ancestry == 'EUR'), EAS = any(ancestry == 'EAS'),
                     min_padj = min(p.adjust),
                     modalities = paste(sort(unique(modality)), collapse = '; ')),
                 by = .(db, ID, Description)]
contrast[, class := fifelse(EUR & EAS, 'shared', fifelse(EUR, 'EUR-only', 'EAS-only'))]
n_shared <- nrow(contrast[class == 'shared']); n_eur <- nrow(contrast[class == 'EUR-only'])
n_eas <- nrow(contrast[class == 'EAS-only'])
```

Of the `r nrow(contrast)` unique significant terms (HPLRv2 background): **`r n_shared` shared**,
**`r n_eur` EUR-only**, **`r n_eas` EAS-only**.

- **Shared**: the ribosome/cytoplasmic-translation cluster (GO BP/MF/CC + KEGG Ribosome),
  ubiquitin-ligase binding, and the STB syncytialization set.
- **EAS-only**: the MHC-I antigen-processing cluster (GO BP antigen presentation; KEGG antigen
  processing, allograft rejection, GvHD, T1D, autoimmune thyroid disease, viral myocarditis;
  GO CC MHC protein complex / ER membrane), consistent with ancestry-differentiated HLA
  regulation; plus O-methyltransferase activity (EAS splicing).
- **EUR-only**: lysosomal lumen (isoform expression), ECM/cell-adhesion terms (intron retention),
  SWI/SNF chromatin-remodeling complexes (alt_TSS), and both pregnancy-set signals
  (PE GWAS candidates in isoforms/splicing; note STB is shared via EAS alt_polyA).

```{r fig2, fig.width=8.5, fig.height=14}
per_mod <- sigh[, .(class = { e <- any(ancestry == 'EUR'); a <- any(ancestry == 'EAS')
                              fifelse(e & a, 'shared', fifelse(e, 'EUR-only', 'EAS-only')) }[1]),
                by = .(modality, db, ID, Description)]
tile <- per_mod[, .(modality, db, Description, class)]
tile[, term := substr(Description, 1, 50)]
tile[, class := factor(class, levels = c('shared','EUR-only','EAS-only'))]
term_order <- unique(tile[order(class, db, term)]$term)
tile[, term := factor(term, levels = rev(term_order))]
tile[, modality := factor(modality, levels = modalities)]
ggplot(tile, aes(x = modality, y = term, fill = class)) +
  geom_tile(color = 'white', linewidth = 0.3) +
  scale_fill_manual(values = c('shared' = '#7570b3', 'EUR-only' = '#1b9e77', 'EAS-only' = '#d95f02')) +
  labs(x = NULL, y = NULL, fill = 'Ancestry class',
       title = 'Significant terms (FDR < 0.05, HPLRv2 background) by ancestry class') +
  theme_bw(base_family = 'Liberation Sans') +
  theme(axis.text.y = element_text(size = 6),
        axis.text.x = element_text(size = 8, angle = 30, hjust = 1),
        plot.title = element_text(size = 10, face = 'bold'),
        legend.text = element_text(size = 8), legend.title = element_text(size = 8),
        panel.grid = element_blank())
```

## Background sensitivity

```{r background-sensitivity}
sig_wide <- dcast(sig[, .(modality, ancestry, db, ID, Description, background)],
                  modality + ancestry + db + ID + Description ~ background,
                  value.var = 'background', fun.aggregate = length)
for (cl in c('hplrv2','tested')) if (!cl %in% names(sig_wide)) sig_wide[, (cl) := 0L]
sig_wide[, concord := fifelse(hplrv2 > 0 & tested > 0, 'both',
                       fifelse(hplrv2 > 0, 'hplrv2-only', 'tested-only'))]
conc_tab <- sig_wide[, .N, by = concord]
knitr::kable(conc_tab, caption = 'Concordance of significant term-strata between backgrounds')
```

`r conc_tab[concord=='both']$N` significant term-strata are detected under both backgrounds,
`r conc_tab[concord=='hplrv2-only']$N` only with HPLRv2, and
`r conc_tab[concord=='tested-only']$N` only with the tested-genes background. All headline
signals (ribosome/translation in alt_polyA, MHC-I in EAS isoforms, STB in EAS alt_polyA,
PE GWAS in EUR isoforms) are significant under **both** backgrounds. The tested-only gains are
dominated by transcription-factor activity terms in splicing strata, where the narrower tested
universe increases power for that class; the HPLRv2-only losses are mostly ribosomal/secretory
terms whose background frequency is higher in the full reference. Conclusions are robust to
background choice.

# Limitations

1. **Large-set dilution.** The biggest strata (EUR splicing: 5,036 significant genes = 40% of the
   annotated universe; isoform expression: ~1,180 genes) approach a large fraction of the
   background, so ORA has little power to detect specific biology there — e.g., canonical
   "spliceosome" terms are *not* enriched in splicing strata (FDR = 1). Significant terms in
   large strata should be read as the strongest deviations, not the only biology present.
2. **Annotation visibility.** `r n_novel_ref` novel HPLRG_ genes and Ensembl genes missing from
   org.Hs.eg.db (1,260 of `r format(n_hplr, big.mark=',')`) are invisible to GO/KEGG ORA.
3. **KEGG immune pathways** (allograft rejection, GvHD, T1D, etc.) are overlapping HLA gene sets;
   they represent one signal, not six independent ones.
4. **Small curated sets.** PE GWAS (13 genes in universe) and WP iron metabolism (9) have limited
   power; WP is only testable in strata where >= 3 of its genes are in the universe.
5. **ORA is not causal.** Enrichment of PE GWAS candidates among xQTL genes is consistent with —
   but does not prove — regulatory mediation of those GWAS signals; colocalization is the
   appropriate follow-up.
6. KEGG results depend on the live KEGG API snapshot at run time.

# Methods summary

- **Input**: ungrouped top-variant cis-QTL scan (tensorQTL) per modality x ancestry; significant
  genes = unique genes with q < 0.05 among tested features.
- **Universes**: HPLRv2.0 PANTRY GTF gene set (primary) or per-stratum tested genes (sensitivity).
- **Tests**: `clusterProfiler` 4.14 `enrichGO` (ENSEMBL keys), `enrichKEGG` (Entrez via `bitr`),
  `enricher` (Hallmark via `msigdbr`; custom pregnancy collection). BH FDR within each
  stratum x database x background; significance at FDR < 0.05. Pregnancy sets tested with
  `minGSSize = 3`; all others use clusterProfiler defaults.
- **Reproducibility**: gene sets are rebuilt from scan files at render time; pregnancy sets and
  ORA results are loaded from caches built by `enrichment_pipeline.R` (full ORA run ~25 min).

# Session info

```{r session-info}
sessionInfo()
```
